Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-29 sources/bmc_template_submission/bmc_article_submission_2021-01-08.tex
fixing quotes and some citations

Assembling a DNA alignment from such a massive database as GenBank can be done "by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been developed and used to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper, pearse2013phylogenerator}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

@@

Physcraper is implemented as a Python pipeline that uses OpenTree's programmatic access
protocols (API's) to automatically
link any phylogeny mapped to OpenTree's standardized taxonomy \cite{mctavish2015phylesystem},
to alignments
from TreeBASE \cite{sanderson1994treebase}, and data from GenBank \cite{benson2000genbank}.
Its utility and functionalities are presented with a case-study analysis of a
group of flowering plants, the hollies.

@@

The Basic Local Alignment Search Tool, BLAST \cite{altschul1990basic} is used for DNA
sequence search on a remote or local GenBank database. It is constrained to a
"search taxon", a taxonomic group in the NCBI taxonomy that is automatically
identified using the OpenTree's taxonomic Most Recent Common Ancestor (MRCA) API
\cite{mrcaAPI, rees2017automated}, as the MRCA of all ingroup taxa that is
also a named clade in the NCBI taxonomy (Figure 1).

BLAST is performed using the `blastn` algorithm \cite{camacho2009blast}
implemented in BioPython's \cite{cock2009biopython} NCBIWWW module \cite{ncbiwww}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excluded for analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a min
and max length threshold, defined as the proportion of the average length without
gaps of all sequences in input alignment (default values of 80% and 120%,
respectively); 4) or if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in OpenTree or
NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed
(default to 5).

@@

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as "hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). The tree sampling 41 species is also available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

@@

\par{\textbf{Project name:} Physcraper}

@@

\par{\textbf{OpenTree:} The Open Tree of Life project}

@@

The datasets generated and analysed during the current study are available in
the repositories "physcraper" containing the source code,
\href{https://github.com/McTavishLab/physcraper}{https://github.com/McTavishLab/physcraper};
"physcraperex" containing the examples,
\href{https://github.com/McTavishLab/physcraperex}{https://github.com/McTavishLab/physcraperex};
and, "physcraper\_ms" containing this reproducible manuscript,
\href{https://github.com/McTavishLab/physcraper\_ms}{https://github.com/McTavishLab/physcraper\_ms}.

@@

\section*{Funding}
This research was supported by the grant "Sustaining the Open Tree of Life", NSF ABI
No. 1759838, and ABI No. 1759846.
Computer time was provided by the Multi-Environment Research Computer for
Exploration and Discovery (MERCED) cluster from the University of California,
Merced (UCM), supported by the NSF Grant No. ACI-1429783.

\section*{Authors' contributions}
LLSR wrote manuscript, alignment code, documentation, performed analyses and developed examples; MK wrote code for ncbidataparser module, filtering of sequences per OTU and using offline blast searches, wrote documentation and tests; EJM conceived study, wrote most of the code, documentation and tests.
All authors contributed to the manuscript and gave final approval for publication.
\ldots

\section*{Acknowledgements}
We thank the members of the OpenTree development team and the "short bar" Science
and Engineering Building 1, UCM, joint lab paper discussion group for valuable
comments on this manuscript.

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper from original
  \cite{gottlieb2005molecular} tree in B.
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and red, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences in the original alignment that generated
  tree in B), filtered them following criteria from section "DNA sequence search and
  filtering", aligned them to original alignment using MUSCLE and performed a
  phylogenetic reconstruction using RAxML with 100 bootstraps. B-D conflict analyses
  performed with OpenTree tools.}
\end{figure}

>>>>

Assembling a DNA alignment from such a massive database as GenBank can be done ``by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been developed and used to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper, pearse2013phylogenerator}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

@@

Physcraper is implemented as a Python pipeline that uses OpenTree's programmatic access
protocols (API's) to automatically
link any phylogeny mapped to OpenTree's standardized taxonomy \cite{mctavish2015phylesystem},
to alignments
from TreeBASE, and data from GenBank.
Its utility and functionalities are presented with a case-study analysis of a
group of flowering plants, the hollies.

@@

The Basic Local Alignment Search Tool, BLAST \cite{altschul1990basic} is used for DNA
sequence search on a remote or local GenBank database. It is constrained to a
 ``search taxon", a taxonomic group in the NCBI taxonomy that is automatically
identified using the OpenTree's taxonomic Most Recent Common Ancestor (MRCA) API
\cite{mrcaAPI, rees2017automated}, as the MRCA of all ingroup taxa that is
also a named clade in the NCBI taxonomy (Figure 1).

BLAST is performed using the blastn algorithm \cite{camacho2009blast}
implemented in BioPython's \cite{cock2009biopython} NCBIWWW module \cite{ncbiwww}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excluded for analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a min
and max length threshold, defined as the proportion of the average length without
gaps of all sequences in input alignment (default values of 80% and 120%,
respectively); 4) or if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in OpenTree or
NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed
(default to 5).

@@

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as ``hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

An online literature review in June 2020 (Google scholar search for ``ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). The tree sampling 41 species is also available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

@@

\textbf{Project name:} Physcraper

@@

\textbf{OpenTree:} The Open Tree of Life project

@@

The datasets generated and analysed during the current study are available in
the repositories ``physcraper" containing the source code,
\href{https://github.com/McTavishLab/physcraper}{https://github.com/McTavishLab/physcraper};
 ``physcraperex" containing the examples,
\href{https://github.com/McTavishLab/physcraperex}{https://github.com/McTavishLab/physcraperex};
and, ``physcraper\_ms" containing this reproducible manuscript,
\href{https://github.com/McTavishLab/physcraper\_ms}{https://github.com/McTavishLab/physcraper\_ms}.

@@

\section*{Funding}
This research was supported by the grant ``Sustaining the Open Tree of Life", NSF ABI
No. 1759838, and ABI No. 1759846.
Computer time was provided by the Multi-Environment Research Computer for
Exploration and Discovery (MERCED) cluster from the University of California,
Merced (UCM), supported by the NSF Grant No. ACI-1429783.

\section*{Authors' contributions}
LLSR wrote manuscript, alignment code, documentation, performed analyses and developed examples; MK wrote code for ncbidataparser module, filtering of sequences per OTU and using offline blast searches, wrote documentation and tests; EJM conceived study, wrote most of the code, documentation and tests.
All authors contributed to the manuscript and gave final approval for publication.

\section*{Acknowledgements}
We thank the members of the OpenTree development team and the ``short bar" Science
and Engineering Building 1, UCM, joint lab paper discussion group for valuable
comments on this manuscript.

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper from original
  \cite{gottlieb2005molecular} tree in B.
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and red, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences in the original alignment that generated
  tree in B), filtered them following criteria from section ``DNA sequence search and
  filtering", aligned them to original alignment using MUSCLE and performed a
  phylogenetic reconstruction using RAxML with 100 bootstraps. B-D conflict analyses
  performed with OpenTree tools.}
\end{figure}
